---
title: "ST2195 – Epsilon-Greedy Multi-Armed Bandit (R)"
author: "Hasan Rashed Mohamed Almadani — Student ID: stu-00104554"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

# Introduction
This report details the implementation and analysis of the Epsilon-Greedy algorithm for the Multi-Armed Bandit problem. Our goal is to find the best of five advertising "arms" by balancing exploration of new options and exploitation of the best-known option. This document provides solutions for Part A (single run simulation), Part B (parameter analysis), and Part C (optimistic initialization) of the assignment, including visualisations and a comprehensive analysis of the results.

## Student Information
- **Name:** Hasan Rashed Mohamed Almadani
- **Student ID:** stu-00104554

```{r setup, include=FALSE}
# Global options and reproducibility
knitr::opts_chunk$set(fig.width = 9, fig.height = 6, fig.align = 'center',
                      echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2025)
```

## 1) Configuration
This section defines all the key parameters for our experiments, centralizing them for easy modification and clarity.

```{r config}
# --- General Parameters ---
N_STEPS <- 10000 # Number of iterations/visitors
RANDOM_SEED <- 2025 # For reproducibility

# --- Bandit Parameters ---
# The true probability of success for each of the K=5 arms
TRUE_PROBS <- c(0.10, 0.50, 0.20, 0.05, 0.30)

# --- Experiment Parameters ---
# Epsilon values to test in Part B
EPSILON_VALUES <- c(0.01, 0.10, 0.50)

# Balanced epsilon for Parts A & C
BALANCED_EPSILON <- 0.10

# Optimistic Q-value for Part C
OPTIMISTIC_Q <- 5.0
```

## 2) Core Functions
Here, we define the essential functions for the Epsilon-Greedy algorithm, cumulative average calculation, and plotting. The `run_bandit` function has been extended to accept an `initial_q` value and to also return instantaneous `regrets`. A `plot_regret` function has also been added.

```{r core-functions}
# Epsilon-Greedy Bandit implementation
run_bandit <- function(epsilon, n_steps, true_probs, initial_q = 0.0) {
  K <- length(true_probs)
  Q <- rep(initial_q, K) # Initialize Q-values with initial_q
  N_arm <- rep(0L, K)
  rewards <- numeric(n_steps)
  choices <- integer(n_steps)
  regrets <- numeric(n_steps) # To store instantaneous regret

  p_opt <- max(true_probs) # The true probability of the optimal arm

  for (i in seq_len(n_steps)) {
    # Explore vs Exploit
    if (runif(1) < epsilon) {
      arm <- sample.int(K, 1)
    } else {
      arm <- which.max(Q)
    }

    # Bernoulli reward for chosen arm
    reward <- if (runif(1) < true_probs[arm]) 1.0 else 0.0

    # Calculate instantaneous regret
    regrets[i] <- p_opt - true_probs[arm]

    # Incremental mean update for the chosen arm
    N_arm[arm] <- N_arm[arm] + 1L
    Q[arm] <- Q[arm] + (reward - Q[arm]) / N_arm[arm]

    rewards[i] <- reward
    choices[i] <- arm
  }

  list(Q = Q, N_arm = N_arm, rewards = rewards, choices = choices, regrets = regrets)
}

cumulative_average <- function(x) {
  cumsum(x) / seq_along(x)
}

plot_cumavg <- function(curves, labels, p_opt = NULL,
                        title = "Cumulative Average Reward",
                        colors = NULL) {
  if (is.null(colors)) {
    colors <- c("#1f77b4", "#2ca02c", "#d62728", "#9467bd", "#8c564b")
  }
  n <- length(curves)
  if (n == 0) stop("curves cannot be empty")

  plot(curves[[1]], type = "l", col = colors[1], lwd = 2,
       xlab = "Iteration", ylab = "Cumulative average reward", main = title,
       ylim = c(min(unlist(curves), na.rm = TRUE), max(unlist(curves), na.rm = TRUE, p_opt)))
  if (n > 1) {
    for (j in 2:n) lines(curves[[j]], col = colors[j], lwd = 2)
  }
  if (!is.null(p_opt)) {
    abline(h = p_opt, col = "black", lty = 2, lwd = 2)
    legend("bottomright", legend = c(labels, sprintf("Theoretical max = %.2f", p_opt)),
           col = c(colors[seq_len(n)], "black"), lty = c(rep(1, n), 2), lwd = 2, bty = "n")
  } else {
    legend("bottomright", legend = labels, col = colors[seq_len(n)], lty = 1, lwd = 2, bty = "n")
  }
}

plot_regret <- function(curves, labels, title = "Cumulative Regret", colors = NULL) {
  if (is.null(colors)) {
    colors <- c("#1f77b4", "#2ca02c", "#d62728", "#9467bd", "#8c564b")
  }
  n <- length(curves)
  if (n == 0) stop("curves cannot be empty")

  plot(curves[[1]], type = "l", col = colors[1], lwd = 2,
       xlab = "Iteration", ylab = "Cumulative Regret", main = title,
       ylim = c(0, max(unlist(curves), na.rm = TRUE)))
  if (n > 1) {
    for (j in 2:n) lines(curves[[j]], col = colors[j], lwd = 2)
  }
  legend("topleft", legend = labels, col = colors[seq_len(n)], lty = 1, lwd = 2, bty = "n")
}
```

## 3) Part A — Single run with ε = 0.10
We run the simulation for `N_STEPS` iterations with `BALANCED_EPSILON` to observe the algorithm's learning process. Both the cumulative average reward and the cumulative regret are plotted.

```{r part-a}
res_A <- run_bandit(BALANCED_EPSILON, N_STEPS, TRUE_PROBS)
cumavg_A <- cumulative_average(res_A$rewards)
cum_regret_A <- cumsum(res_A$regrets)

p_opt <- max(TRUE_PROBS)

# Plot Cumulative Average Reward
plot_cumavg(list(cumavg_A), labels = list(sprintf("epsilon = %.2f", BALANCED_EPSILON)),
            p_opt = p_opt,
            title = "Part A: Cumulative Average Reward")

# Plot Cumulative Regret
plot_regret(list(cum_regret_A), labels = list(sprintf("epsilon = %.2f", BALANCED_EPSILON)),
            title = "Part A: Cumulative Regret")

cat("\nFinal estimates Q:\n"); print(round(res_A$Q, 3))
cat("Arm pulls:\n"); print(res_A$N_arm)
cat("Optimal arm index:", which.max(TRUE_PROBS), "with p=", p_opt, "\n")
```

## 4) Part B — Parameter Analysis (Explore–Exploit Trade-off)
Here, we compare the algorithm's performance across different epsilon values (`EPSILON_VALUES`) to understand the exploration-exploitation trade-off.

```{r part-b}
results <- list()
cumavg_curves <- list(); cumavg_labels <- list()
regret_curves <- list(); regret_labels <- list()

for (eps in EPSILON_VALUES) {
  res <- run_bandit(eps, N_STEPS, TRUE_PROBS)
  results[[as.character(eps)]] <- res # Store full results for metrics calculation
  
  cumavg_curves[[length(cumavg_curves) + 1]] <- cumulative_average(res$rewards)
  cumavg_labels[[length(cumavg_labels) + 1]] <- sprintf("epsilon = %.2f", eps)
  
  regret_curves[[length(regret_curves) + 1]] <- cumsum(res$regrets)
  regret_labels[[length(regret_labels) + 1]] <- sprintf("epsilon = %.2f", eps)
}

p_opt <- max(TRUE_PROBS)

# Plot overlay for Cumulative Average Reward
plot_cumavg(cumavg_curves, cumavg_labels, p_opt = p_opt,
            title = "Part B: Cumulative Average Reward vs. epsilon")

# Plot overlay for Cumulative Regret
plot_regret(regret_curves, regret_labels, title = "Part B: Cumulative Regret vs. epsilon")

# ---
# Metrics
# ---
threshold <- 0.95 * p_opt
TAIL <- 1000
mean_p <- mean(TRUE_PROBS)

cat("Convergence (first step to reach >= 95% of optimal):\n")
for (eps in EPSILON_VALUES) {
  y <- cumulative_average(results[[as.character(eps)]]$rewards)
  idx <- which(y >= threshold)[1]
  val <- if (is.na(idx)) "never" else as.character(idx)
  cat(sprintf("  epsilon=%.2f: %s\n", eps, val))
}

cat("\nStability (variance of last 1000 points; smaller is stabler):\n")
for (eps in EPSILON_VALUES) {
  y <- cumulative_average(results[[as.character(eps)]]$rewards)
  tail_y <- tail(y, TAIL)
  cat(sprintf("  epsilon=%.2f: %.6f\n", eps, var(tail_y)))
}

cat("\nAsymptotic expected average reward (approx):\n")
for (eps in EPSILON_VALUES) {
  asym <- (1 - eps) * p_opt + eps * mean_p
  cat(sprintf("  epsilon=%.2f: %.3f\n", eps, asym))
}
```

## 5) Part C — Optimistic Initial Values
This experiment explores **Optimistic Initial Values**, a technique to encourage exploration. Instead of initializing Q-values to 0, we set them to a high, 'optimistic' value (`OPTIMISTIC_Q`). The algorithm is 'disappointed' by the initial, lower rewards, which drives it to explore all other arms at least once before settling on the best one. We compare this to the standard approach using a balanced epsilon.

```{r part-c}
# Run with standard (zero) initialization for comparison
res_std <- run_bandit(BALANCED_EPSILON, N_STEPS, TRUE_PROBS, initial_q = 0.0)
cumavg_std <- cumulative_average(res_std$rewards)
cum_regret_std <- cumsum(res_std$regrets)

# Run with optimistic initialization
res_opt <- run_bandit(BALANCED_EPSILON, N_STEPS, TRUE_PROBS, initial_q = OPTIMISTIC_Q)
cumavg_opt <- cumulative_average(res_opt$rewards)
cum_regret_opt <- cumsum(res_opt$regrets)

p_opt <- max(TRUE_PROBS)

# Plot Cumulative Average Reward
plot_cumavg(list(cumavg_std, cumavg_opt),
            labels = list("Standard Init (Q=0)", sprintf("Optimistic Init (Q=%.1f)", OPTIMISTIC_Q)),
            p_opt = p_opt,
            title = sprintf("Part C: Optimistic vs. Standard Initialization (epsilon=%.2f)", BALANCED_EPSILON))

# Plot Cumulative Regret
plot_regret(list(cum_regret_std, cum_regret_opt),
            labels = list("Standard Init (Q=0)", sprintf("Optimistic Init (Q=%.1f)", OPTIMISTIC_Q)),
            title = "Part C: Cumulative Regret Comparison")
```

## 6) Additional Analysis: Arm Pull Distribution
Finally, we can visualize the *behavior* of the standard vs. optimistic agents by plotting how many times they pulled each arm. This gives a clear picture of their exploration strategy.

```{r arm-pulls}
# Prepare data for plotting
arm_names <- paste("Arm", 1:length(TRUE_PROBS))
counts_df <- data.frame(
  Arm = rep(arm_names, 2),
  Strategy = c(rep("Standard Init", length(TRUE_PROBS)), rep("Optimistic Init", length(TRUE_PROBS))),
  Pulls = c(res_std$N_arm, res_opt$N_arm)
)

# Plotting using base R barplot
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2) + 0.1) # Set up 2 plots side-by-side
barplot(res_std$N_arm, names.arg = arm_names, main = "Standard Init: Arm Pulls",
        ylab = "Number of Pulls", xlab = "Arm", col = "skyblue")
barplot(res_opt$N_arm, names.arg = arm_names, main = "Optimistic Init: Arm Pulls",
        ylab = "Number of Pulls", xlab = "Arm", col = "lightcoral")
par(mfrow = c(1, 1)) # Reset plot layout
```

## 7) Analysis & Conclusion
Based on the experiments conducted, we can draw the following conclusions regarding the Epsilon-Greedy algorithm's behavior:

1.  **The Explore-Exploit Tradeoff (Part B)**: Our analysis confirms that choosing an appropriate epsilon value is crucial. A very low `epsilon` (0.01), while appearing "greedy," was slower to converge on the optimal arm but showed high long-term stability once it did. Conversely, a high `epsilon` (0.50) led to excessive exploration, preventing the algorithm from fully exploiting the best arm and thus keeping the cumulative reward significantly below the theoretical maximum. The cumulative regret plots clearly illustrate that a higher epsilon leads to a greater total regret, signifying the cost of over-exploration.

2.  **Optimistic Initialization (Part C)**: This experiment revealed the effectiveness of optimistic initialization as a strategy for forced exploration. By starting with high initial Q-values (5.0), the agent was compelled to try all arms, as initial rewards would invariably be lower than its optimistic belief, driving it to explore other seemingly better options. This is evident in the arm pull distribution, where the optimistic agent explored all arms more evenly compared to the standard agent which quickly focused on the highest-reward arm. While for this specific problem `epsilon=0.1` without optimistic initialization proved efficient, optimistic initialization can be invaluable in more complex scenarios to prevent premature convergence to suboptimal solutions. The initial high regret for the optimistic agent highlights the "cost" of this forced initial exploration.