---
title: "ST2195 – Epsilon-Greedy Multi-Armed Bandit (R)"
author: "Hasan Rashed Mohamed Almadani — Student ID: stu-00104554"
date: "15/12/2025"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

# Introduction
This report implements and analyzes the Epsilon-Greedy algorithm for the Multi-Armed Bandit problem. It aims to find the best of five advertising "arms" by balancing exploration and exploitation. Solutions for Part A (simulation), Part B (parameter analysis), and Part C (optimistic initialization) are provided, with visualizations and analysis.

## Student Information
- **Name:** Hasan Rashed Mohamed Almadani
- **Student ID:** stu-00104554

```{r setup, include=FALSE}
# Global options and reproducibility
knitr::opts_chunk$set(fig.width = 9, fig.height = 6, fig.align = 'center',
                      echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2025)
```

## 1) Configuration
This section defines all the key parameters for our experiments, centralizing them for easy modification and clarity.

```{r config}
# --- General Parameters ---
N_STEPS <- 10000 # Number of iterations
RANDOM_SEED <- 2025 # For reproducibility

# --- Bandit Parameters ---
# The true probability of success for each of the K=5 arms
TRUE_PROBS <- c(0.10, 0.50, 0.20, 0.05, 0.30)

# --- Experiment Parameters ---
# Epsilon values to test in Part B
EPSILON_VALUES <- c(0.01, 0.10, 0.50)

# Balanced epsilon for Parts A & C
BALANCED_EPSILON <- 0.10

# Optimistic Q-value for Part C
OPTIMISTIC_Q <- 5.0
```

## 2) Core Functions
Defines functions for the Epsilon-Greedy algorithm, cumulative average, and plotting. The `run_bandit` function now includes `initial_q` and returns `regrets`. A `plot_regret` function is also included.

```{r core-functions}
# Epsilon-Greedy Bandit implementation
run_bandit <- function(epsilon, n_steps, true_probs, initial_q = 0.0) {
  K <- length(true_probs)
  Q <- rep(initial_q, K) # Initializes Q-values
  N_arm <- rep(0L, K)
  rewards <- numeric(n_steps)
  choices <- integer(n_steps)
  regrets <- numeric(n_steps) # Stores instantaneous regrets

  p_opt <- max(true_probs) # The true probability of the optimal arm

  for (i in seq_len(n_steps)) {
    # Explore vs Exploit
    if (runif(1) < epsilon) {
      arm <- sample.int(K, 1)
    } else {
      arm <- which.max(Q)
    }

    # Bernoulli reward
    reward <- if (runif(1) < true_probs[arm]) 1.0 else 0.0

    # Calculates instantaneous regrets
    regrets[i] <- p_opt - true_probs[arm]

    # Update incremental mean
    N_arm[arm] <- N_arm[arm] + 1L
    Q[arm] <- Q[arm] + (reward - Q[arm]) / N_arm[arm]

    rewards[i] <- reward
    choices[i] <- arm
  }

  list(Q = Q, N_arm = N_arm, rewards = rewards, choices = choices, regrets = regrets)
}

cumulative_average <- function(x) {
  cumsum(x) / seq_along(x)
}

plot_cumavg <- function(curves, labels, p_opt = NULL,
                        title = "Cumulative Average Reward",
                        colors = NULL) {
  if (is.null(colors)) {
    colors <- c("blue", "green", "red", "purple", "brown")
  }
  n <- length(curves)
  if (n == 0) stop("curves cannot be empty")

  plot(curves[[1]], type = "l", col = colors[1], lwd = 2,
       xlab = "Iteration", ylab = "Cumulative average reward", main = title,
       ylim = c(min(unlist(curves), na.rm = TRUE), max(unlist(curves), na.rm = TRUE, p_opt)))
  if (n > 1) {
    for (j in 2:n) lines(curves[[j]], col = colors[j], lwd = 2)
  }
  if (!is.null(p_opt)) {
    abline(h = p_opt, col = "black", lty = 2, lwd = 2)
    legend("bottomright", legend = c(labels, sprintf("Theoretical max = %.2f", p_opt)),
           col = c(colors[seq_len(n)], "black"), lty = c(rep(1, n), 2), lwd = 2, bty = "n")
  } else {
    legend("bottomright", legend = labels, col = colors[seq_len(n)], lty = 1, lwd = 2, bty = "n")
  }
}

plot_regret <- function(curves, labels, title = "Cumulative Regret", colors = NULL) {
  if (is.null(colors)) {
    colors <- c("blue", "green", "red", "purple", "brown")
  }
  n <- length(curves)
  if (n == 0) stop("curves cannot be empty")

  plot(curves[[1]], type = "l", col = colors[1], lwd = 2,
       xlab = "Iteration", ylab = "Cumulative Regret", main = title,
       ylim = c(0, max(unlist(curves), na.rm = TRUE)))
  if (n > 1) {
    for (j in 2:n) lines(curves[[j]], col = colors[j], lwd = 2)
  }
  legend("topleft", legend = labels, col = colors[seq_len(n)], lty = 1, lwd = 2, bty = "n")
}
```

## 3) Part A — Single run with ε = 0.10
Simulation runs for `N_STEPS` iterations with `BALANCED_EPSILON`. Both cumulative average reward and regret are plotted.

```{r part-a}
res_A <- run_bandit(BALANCED_EPSILON, N_STEPS, TRUE_PROBS)
cumavg_A <- cumulative_average(res_A$rewards)
cum_regret_A <- cumsum(res_A$regrets)

p_opt <- max(TRUE_PROBS)

# Plot the Cumulative Average Reward
plot_cumavg(list(cumavg_A), labels = list(sprintf("epsilon = %.2f", BALANCED_EPSILON)),
            p_opt = p_opt,
            title = "Part A: Cumulative Average Reward")

# Plot the Cumulative Regret
plot_regret(list(cum_regret_A), labels = list(sprintf("epsilon = %.2f", BALANCED_EPSILON)),
            title = "Part A: Cumulative Regret")

cat("\nFinal estimates Q:\n"); print(round(res_A$Q, 3))
cat("Arm pulls:\n"); print(res_A$N_arm)
cat("Optimal arm index:", which.max(TRUE_PROBS), "with p=", p_opt, "\n")
```

## 4) Part B — Parameter Analysis
Compares algorithm performance across `EPSILON_VALUES` to study the exploration-exploitation trade-off.

```{r part-b}
results <- list()
cumavg_curves <- list(); cumavg_labels <- list()
regret_curves <- list(); regret_labels <- list()

for (eps in EPSILON_VALUES) {
  res <- run_bandit(eps, N_STEPS, TRUE_PROBS)
  results[[as.character(eps)]] <- res # Store full results for metrics calculation
  
  cumavg_curves[[length(cumavg_curves) + 1]] <- cumulative_average(res$rewards)
  cumavg_labels[[length(cumavg_labels) + 1]] <- sprintf("epsilon = %.2f", eps)
  
  regret_curves[[length(regret_curves) + 1]] <- cumsum(res$regrets)
  regret_labels[[length(regret_labels) + 1]] <- sprintf("epsilon = %.2f", eps)
}

p_opt <- max(TRUE_PROBS)

# Plot overlay for Cumulative Average Reward
plot_cumavg(cumavg_curves, cumavg_labels, p_opt = p_opt,
            title = "Part B: Cumulative Average Reward vs. epsilon")

# Plot overlay for Cumulative Regret
plot_regret(regret_curves, regret_labels, title = "Part B: Cumulative Regret vs. epsilon")

# ---
# Metrics
# ---
threshold <- 0.95 * p_opt
TAIL <- 1000
mean_p <- mean(TRUE_PROBS)

cat("Convergence (first step to reach >= 95% of optimal):\n")
for (eps in EPSILON_VALUES) {
  y <- cumulative_average(results[[as.character(eps)]]$rewards)
  idx <- which(y >= threshold)[1]
  val <- if (is.na(idx)) "never" else as.character(idx)
  cat(sprintf("  epsilon=%.2f: %s\n", eps, val))
}

cat("\nStability (variance of last 1000 points; smaller is stabler):\n")
for (eps in EPSILON_VALUES) {
  y <- cumulative_average(results[[as.character(eps)]]$rewards)
  tail_y <- tail(y, TAIL)
  cat(sprintf("  epsilon=%.2f: %.6f\n", eps, var(tail_y)))
}

cat("\nAsymptotic expected average reward (approx):\n")
for (eps in EPSILON_VALUES) {
  asym <- (1 - eps) * p_opt + eps * mean_p
  cat(sprintf("  epsilon=%.2f: %.3f\n", eps, asym))
}
```

## 5) Part C — Optimistic Initial Values
Explores **Optimistic Initial Values** for forced exploration. Q-values are set to `OPTIMISTIC_Q` (high), compelling the algorithm to explore all arms. Compared to standard initialization.

```{r part-c}
# Standard (zero) initialization
res_std <- run_bandit(BALANCED_EPSILON, N_STEPS, TRUE_PROBS, initial_q = 0.0)
cumavg_std <- cumulative_average(res_std$rewards)
cum_regret_std <- cumsum(res_std$regrets)

# Optimistic initialization
res_opt <- run_bandit(BALANCED_EPSILON, N_STEPS, TRUE_PROBS, initial_q = OPTIMISTIC_Q)
cumavg_opt <- cumulative_average(res_opt$rewards)
cum_regret_opt <- cumsum(res_opt$regrets)

p_opt <- max(TRUE_PROBS)

# Plot Cumulative Average Reward
plot_cumavg(list(cumavg_std, cumavg_opt),
            labels = list("Standard Init (Q=0)", sprintf("Optimistic Init (Q=%.1f)", OPTIMISTIC_Q)),
            p_opt = p_opt,
            title = sprintf("Part C: Optimistic vs. Standard Initialization (epsilon=%.2f)", BALANCED_EPSILON))

# Plot Cumulative Regret
plot_regret(list(cum_regret_std, cum_regret_opt),
            labels = list("Standard Init (Q=0)", sprintf("Optimistic Init (Q=%.1f)", OPTIMISTIC_Q)),
            title = "Part C: Cumulative Regret Comparison")
```

## 6) Additional Analysis: Arm Pull Distribution
Visualizes arm pull distribution for standard vs. optimistic agents to show exploration strategy.

```{r arm-pulls}
# Prepare plotting data
arm_names <- paste("Arm", 1:length(TRUE_PROBS))
counts_df <- data.frame(
  Arm = rep(arm_names, 2),
  Strategy = c(rep("Standard Init", length(TRUE_PROBS)), rep("Optimistic Init", length(TRUE_PROBS))),
  Pulls = c(res_std$N_arm, res_opt$N_arm)
)

# Base R barplot
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2) + 0.1) # Set 2 plots side-by-side
barplot(res_std$N_arm, names.arg = arm_names, main = "Standard Init: Arm Pulls",
        ylab = "Number of Pulls", xlab = "Arm", col = "skyblue")
barplot(res_opt$N_arm, names.arg = arm_names, main = "Optimistic Init: Arm Pulls",
        ylab = "Number of Pulls", xlab = "Arm", col = "lightcoral")
par(mfrow = c(1, 1)) # Reset layout
```

## 7) Analysis & Conclusion
Conclusions on the Epsilon-Greedy algorithm's behavior from conducted experiments:

1.  **The Explore-Exploit Tradeoff (Part B)**: Appropriate epsilon is crucial. Low `epsilon` (0.01) converges slowly but is stable. High `epsilon` (0.50) over-explores, preventing optimal exploitation and leading to higher cumulative regret.

2.  **Optimistic Initialization (Part C)**: Optimistic initialization effectively forces exploration by setting high initial Q-values (5.0). This drives the agent to try all arms, evident in more even arm pull distribution compared to standard methods. Useful in complex scenarios to avoid suboptimal convergence, despite initial high regret.